---
title: "testing2"
format: html
---

# TESTING

## INSIGHTFUL EFFECTIVE TESTING

Good tests require you to *think* through what you want your code to do

- What should the output be for a given input?
- What should the output be for a range of inputs?
- What should the output be for edge cases (e.g. zero, negative, etc.)?

Requires you to come up with situations where you *know* what the answer should be — these make good tests.

*Especially important when a function is part of a larger program*

## EXPECTATIONS

Testing in R uses **expectations**.

Expectations are functions you use to verify that your code is working as intended:

- expect_true(condition)
- expect_equal(value, value)
- expect_match(string1, string2)
- expect_null(value)
- expect_length(object, n)

testthat reference

## TESTING IN R — TEST_THAT

A **test** runs one or more **expectation** functions and reports a message.

You can also add test data to the test

Format: test_that(description, { series of expectations })

```{r}
library(tidyverse)
library(here)
library(devtools)
library(testthat)

source(here("Week6/compute_NPV.R"))
# if profit is 0, NPV should be 0
# if time is 0, NPV should equal the original value

# now lets write the test that combines several expectations
test_that("compute_NPV_works", {

expect_equal(compute_NPV(value=0,time=100,discount=0.1), 0)
expect_equal(compute_NPV(value=100, time=0, discount=0.1), 100)
})

# now lets add a test that checks that NPV is higher for a shorter time period
test_that("compute_NPV_time_works", {
testdata = c(2, 4)
short = compute_NPV(100, time=testdata[1], 0.01)
long = compute_NPV(100, time = testdata[2], 0.01)
expect_true(short > long)
})
```

## TESTING ERROR HANDLING

- expect_warning — checks that a warning is returned
- expect_error — checks that an error is returned
- expect_message — checks that a message is returned
- expect_condition - checks that a condition is returned (more general)
- Warning versus Error*

Use a *warning* when you want function to still return a result, but you want to alert the user that something MIGHT be wrong with the input.

Use a *error* when you want the function to stop and not return a result if the input is wrong. in R, in the function

- use stop() to throw an error
- use warning() to throw a warning
- use message() to throw a message

I’ll illustrate using a different version of compute_NPV ideally you’d fix in your git workflow and commit

```{r}
source(here("course-materials/R/compute_NPV_witherror.R"))
test_that("compute_NPV_haserrorchecking", {
  # throw an error if time is negative 
expect_error(compute_NPV(value=100, time=-1, discount=-0.1))  
  # throw an error if discount is negative
expect_error(compute_NPV(value=100, time=1, discount=-0.1))  
  # throw a warning if value is negative
 expect_warning(compute_NPV(value=-100, time=1, discount=0.1))
})
```

## USING TESTS

What we really want is to run tests *every time* we change something.

Store multiple tests in a *name.R* file and keep those files in the *tests/testthat* directory.

Two options: - *test_dir(“name”)* — runs all tests in the “name” subdirectory (all files beginning with the word “test”) - *test_file(“name”)* — runs all the tests in a file called “name”

**Goal:** Write tests for each function in your package/project.

Organization options:

- All tests for a given function in one file; all test files in one directory (*most common*)
- All test types in one file (*less common*)

## EXAMPLES

- test_compute_NPV.R — tests for compute_NPV function

To illustrate we will try it on both versions so we can see what happens when we add error checking and when we don’t have error checking.

```{r}
# source original function
source(here("Week6/compute_NPV.R"))
# run tests in a file
test_file(here("Week7/tests/test_compute_NPV.R"))

# now try  version with error checking
source(here("course-materials/R/compute_NPV_witherror.R"))
test_file(here("Week7/tests/test_compute_NPV.R"))
```

## ANOTHER EXAMPLE

Lets consider a function designed to extract information of spring temperatures from climate data. We want to make sure it is working correctly and that it is robust to different inputs.

As always, define contract - inputs/outputs

The function takes as input climate data with columns:

- year
- month
- average temperature (tavg)

*Optional Inputs*

- *spring_months* (vector of integers) to include in spring; default 4,5,6 - s
- *spring_out* (default FALSE) set to TRUE to output average spring temperature for all years

**Outputs**

- *Tavg_spring* mean_springT mean spring temperature (C)
- *Tmax_spring* highest spring temperature (C)
- *Tmin_spring* lowest spring temperature (C)
- *warmest_spring* year with highest spring temperature (year)
- *coldest_spring* year with lowest spring temperature (year)

OPTIONAL array of spring temperature means for all years if spring_out set to TRUE

In error checking, it will make use of *stopifnot* to check that the necessary columns are present in the data frame.

```{r}
# source function
source(here("Week7/spring_summary.R"))
spring_summary
```

```{r}
# to test this function we need to create some data, we can
# add days and precip even though function doesn't use them to make sure it doesn't
# mess up the function
clim_data = as.data.frame(cbind(month=c(1:4), day=rep(1, times=4), year=c(2000,2000,2001,2001), 
                                 precip=rep(0, times=4), tavg=c(2,2,2,4)))
spring_summary(clim_data)

spring_summary(clim_data, spring_months=c(1,4))
```

```{r}
# we can embed data directly in a test
# save this to a file called test_spring_summary.R
test_that("spring_summary_works" ,
{clim_data = as.data.frame(cbind(month=c(1:4), day=rep(1, times=4), year=c(2000,2000,2001,2001), 
                                 precip=rep(0, times=4), tavg=c(2,2,2,4)))
expect_equal(spring_summary(clim_data, spring_months=4)$Tavg_spring, 4)
expect_equal(spring_summary(clim_data, spring_months=2)$Tavg_spring, 2)
expect_equal(spring_summary(clim_data, spring_months=c(1:4))$warmest_spring, 2001)

})
```

```{r}
# run a bunch of tests at once

test_file(here("Week7/tests/test_spring_summary.R"))
```

```{r}
# run all tests
test_dir(here("Week7/tests"))
```

## QUIZ

What expectation function would you add to test_spring_summary to check whether is throws an error if one of the columns are missing?

```{r}
library(here)
source(here("Week7/spring_summary.R"))
test_that("spring_summary_error", {
  clim_data = data.frame(
    month = c(1, 2),
    year = c(2000, 2001)
  )
expect_error(spring_summary(clim_data))

})

test_file(here("Week7/tests/test_spring_summary.R"))
```

