---
title: "apis"
format: html
editor: 
  markdown: 
    wrap: 72
---

# GETTING DATA FROM THE WEB - WORKING WITH APIS

## AGENDA FOR TODAY

- What is an API and why should we care?
- How does the web actually work (just enough to be useful)
- Making API requests in R
- Parsing the data you get back
- Examples: air quality data and streamflow

## WHERE DOES DATA COME FROM?

So far we have mostly worked with data that already lives in a file:

```{r}
# reading in data from a file - what you know already
# data <- read_csv("my_data.csv")
```

Or we’ve generated test data

A huge amount of scientific and environmental data lives **online** and
is updated continuously

- Current air quality at monitoring stations
- Real-time streamflow from USGS gauges
- Species occurrence records from citizen scientists
- Weather observations and forecasts
- Satellite-derived vegetation indices

How do we access that data **directly from R**, without manually
downloading files?

## ENTER: THE API

**API** = Application Programming Interface

Think of it as a **standardized way to ask a remote computer for data**

Analogy: imagine a restaurant

- The **kitchen** is a database full of data (you can’t walk in
  yourself)
- The **menu** is the API — it lists exactly what you can ask for and
  how
- The **waiter** is the HTTP request — carries your order and brings
  back the result
- The **meal** is the data returned to you

## HOW IT WORKS (SIMPLIFIED)

You send a **request** to a URL — a web address with your question baked
in

Request format: Every API Request Has 4 Parts

- protocol (communication method usually HTTPS)
- base URL (where the API lives)
- endpoint (specific data you want)
- query parameters (details of your request, like location, date range,
  variables)

*BUT* you don’t build that yourself — you use R functions to construct
the request

The server reads your request and sends back data — usually in a format
called **JSON**

## WHAT IS JSON?

**JSON** (JavaScript Object Notation) is how most APIs return data

It looks a bit like an R list:

{ "city": "Santa Barbara", "temperature": 18.4, "humidity": 72,
"conditions": "partly cloudy", "wind": { "speed": 3.2, "direction":
"WSW" } }

R has great tools to turn this into a usable data frame

## API KEYS

Many APIs require an **API key** — a unique identifier that:

- Tells the server who you are
- Lets providers track and limit usage
- Is usually **free** for research/student use

**Warning**

**Never share your API key or put it directly in your code!**

Store it as an environment variable instead:

```{r}
# in your .Renviron file (run usethis::edit_r_environ() to open it):
# OPENAQ_KEY=your_key_here

# then in your R script, retrieve it safely:
# my_key <- Sys.getenv("OPENAQ_KEY")
```

Some APIs (including our examples) are open and need no key at all

## OUR TOOLKIT

We’ll use two packages:

- **httr2** — for sending HTTP requests (the “waiter”)
- **jsonlite** — for parsing JSON into R objects (decoding the “meal”)

```{r}
# install.packages("httr2")
# install.packages("jsonlite")

library(httr2)
library(jsonlite)
library(tidyverse)
```

## STANDARD API WORKFLOW

- **Find the API** — read the documentation to understand endpoints and
  parameters
- **Build a request** — use *request* and *req_url_query*, functions to
  construct your query
- **Send it** — *req_perform()* to get the response
- **Parse the response** — usually *resp_body_json()* to turn JSON into
  an R, doesn’t always work *resp_content_type()* can help

## EXAMPLE 1: AIR QUALITY

Using the **Open-Meteo Air Quality API** — free, no key needed

## OPEN-METEO AIR QUALITY

Open-Meteo provides free weather and air quality data

You could just use the website and download CSV file and upload into
R…but API makes it easier

## MAKING THE REQUEST

- API endpoint: https://air-quality-api.open-meteo.com/v1/air-quality
- API host site will provide information on structure of available data
  including variables names and formats, and how to specify parameters
  in the request URL
- API documentation: open-meteo.com/en/docs/air-quality-api

Let’s get hourly PM2.5 data for Santa Barbara:

```{r}
# Build the request URL
req <- request("https://air-quality-api.open-meteo.com/v1/air-quality")  %>%
  req_url_query(
    latitude  = 34.42,
    longitude = -119.70,
    hourly    = "pm2_5", 
    past_days = 3       # last 3 days of data
  )

# Send the request and parse the response
resp <- req %>% req_perform()

# what is this
attributes(resp)

str(resp)
```

```{r}
# Convert  to an R list
# the simplifyVector to format NA's usually what you want
aq_raw <- resp %>% resp_body_json(simplifyVector = TRUE)

# What did we get back?
names(aq_raw)
```

```{r}
# Step 2 — What's inside a key?
str(aq_raw$hourly)          # compact overview of structure

length(aq_raw$hourly$pm_2) # how many values?
```

## PARSING THE RESPONSE

The response is a nested list — we need to pull out what we want:

This will use POSIX - date time format that R can understand, and unlist
to pull out the actual values from the list

We will cover POSIX data format in a future class

```{r}
# Extract what you need to a data frame
aq_df <- data.frame(
  # turn time into a POSIX data format
  time  = as.POSIXct(unlist(aq_raw$hourly$time), format = "%Y-%m-%dT%H:%M"),
  pm2_5 = unlist(aq_raw$hourly$pm2_5)
)

head(aq_df)
```

Now it’s a regular R data frame

## VISUALISING PM2.5

```{r}
ggplot(aq_df, aes(x = time, y = pm2_5)) +
  geom_line(colour = "steelblue") +
  geom_hline(yintercept = 12, linetype = "dashed", colour = "orange") +
  annotate("text", x = min(aq_df$time), y = 13.5,
           label = "EPA annual standard (12 µg/m³)",
           hjust = 0, size = 3, colour = "orange") +
  labs(
    title = "PM2.5 Air Quality — Santa Barbara",
    x     = "Time",
    y     = "PM2.5 (µg/m³)"
  ) +
  theme_minimal()
```

## MODIFYING THE REQUEST

- add *ozone*
- add estimates for next 6 days

Both are specified in the API documentation so you can add them to the
request URL

```{r}
# Build the request URL
req <- request("https://air-quality-api.open-meteo.com/v1/air-quality")  %>%
  req_url_query(
    latitude  = 34.42,
    longitude = -119.70,
    hourly    = "pm2_5,ozone",  # we can ask for multiple variables at once, you can't have spaces!!
    past_days = 3,  # last 3 days of data
    forecast_days = 6 # add forecast days
  )

# Send the request and parse the response
resp <- req %>% req_perform()

# what is this
attributes(resp)
```

```{r}
# Convert  to an R list
aq2_raw <- resp %>% resp_body_json(simplifyVector = TRUE)

# What did we get back?
names(aq2_raw)

names(aq2_raw$hourly)
```

```{r}
# Step 2 — What's inside a key?
names(aq_raw$hourly)          # compact overview of structure

# how many values?
length(aq2_raw$hourly$pm2_5) 
```

## PARSING THE RESPONSE

The response is a nested list — we need to pull out what we want:

```{r}
# Extract what you need to a data frame
aq_df2 <- data.frame(
  # turn time into a POSIX data format
  time  = as.POSIXct(unlist(aq2_raw$hourly$time), format = "%Y-%m-%dT%H:%M"),
  pm2_5 = unlist(aq2_raw$hourly$pm2_5),
  # add ozone to the data frame
  ozone = unlist(aq2_raw$hourly$ozone)
)

head(aq_df2)
```

Now it’s a regular R data frame

## VISUALISING PM2.5 VS OZONE

```{r}
ggplot(aq_df2, aes(x = ozone, y = pm2_5)) +
  geom_point(colour = "steelblue") +
  labs(
    title = "Ozone vs PM2.5 Air Quality — Santa Barbara",
    x     = "Ozone (µg/m³)",
    y     = "PM2.5 (µg/m³)"
  ) +
  theme_minimal()
```

## TRY ON YOUR OWN

Repeat above and extract PM10 instead of PM2.5 Compute mean PM10, and
PM2.5 for your data frame Add to CANVAS quiz

```{r}
# Build the request URL
req2 <- request("https://air-quality-api.open-meteo.com/v1/air-quality")  %>%
  req_url_query(
    latitude  = 34.42,
    longitude = -119.70,
    hourly    = "pm10", 
    past_days = 3       # last 3 days of data
  )

# Send the request and parse the response
resp2 <- req2 %>% req_perform()

# what is this
attributes(resp)

str(resp)
```

```{r}
# Convert  to an R list
# the simplifyVector to format NA's usually what you want
aq3_raw <- resp2 %>% resp_body_json(simplifyVector = TRUE)

# What did we get back?
names(aq3_raw)
```

```{r}
# Step 2 — What's inside a key?
str(aq3_raw$hourly)          # compact overview of structure

length(aq3_raw$hourly$pm10) # how many values?
```

```{r}
# Extract what you need to a data frame
aq2_df <- data.frame(
  # turn time into a POSIX data format
  time  = as.POSIXct(unlist(aq_raw$hourly$time), format = "%Y-%m-%dT%H:%M"),
  pm10 = unlist(aq3_raw$hourly$pm10)
)

head(aq2_df)
```

```{r}
# Build the request URL
req2 <- request("https://air-quality-api.open-meteo.com/v1/air-quality")  %>%
  req_url_query(
    latitude  = 34.42,
    longitude = -119.70,
    hourly    = "pm10,ozone",  # we can ask for multiple variables at once, you can't have spaces!!
    past_days = 3,  # last 3 days of data
    forecast_days = 6 # add forecase days
  )

# Send the request and parse the response
resp2 <- req2 %>% req_perform()

# what is this
attributes(resp2)
```

```{r}
# Convert  to an R list
aq3_raw <- resp2 %>% resp_body_json(simplifyVector = TRUE)

# What did we get back?
names(aq3_raw)

names(aq3_raw$hourly)
```

```{r}
# Step 2 — What's inside a key?
names(aq3_raw$hourly) 
```

```{r}
# Extract what you need to a data frame
aq_df3 <- data.frame(
  # turn time into a POSIX data format
  time  = as.POSIXct(unlist(aq3_raw$hourly$time), format = "%Y-%m-%dT%H:%M"),
  pm10 = unlist(aq3_raw$hourly$pm10),
  # add ozone to the data frame
  ozone = unlist(aq3_raw$hourly$ozone)
)

head(aq_df3)

```

## WRITING A REUSABLE API FUNCTION

Connecting APIs to what we’ve been learning about modularity

## WHY WRITE A FUNCTION?

What if we want air quality data for **many** locations?

This is exactly the modularity pattern from last week — write a function
with a clear **contract**:

*Input*: latitude, longitude

*Output*: a tidy data frame of hourly PM2.5 readings

*Parameters*: number of past days

```{r}
get_pm25 <- function(lat, lon, past_days = 7) {
  
  # Error checking
  if (lat < -90 | lat > 90) {
    return("Latitude must be between -90 and 90.")
  }
  if (lon < -180 | lon > 180) {
    return("Longitude must be between -180 and 180.")
  }                                                    # ← added this
  
  req <- request("https://air-quality-api.open-meteo.com/v1/air-quality") |>
    req_url_query(
      latitude  = lat,
      longitude = lon,
      hourly    = "pm2_5",
      past_days = past_days
    )
  resp <- req |> req_perform()
  raw  <- resp |> resp_body_json(simplifyVector = TRUE)
  data.frame(
    lat   = lat,
    lon   = lon,
    time  = as.POSIXct(unlist(raw$hourly$time), format = "%Y-%m-%dT%H:%M"),
    pm2_5 = unlist(raw$hourly$pm2_5)
  )
}
```

## USING OUR FUNCTION WITH MULTIPLE LOCATIONS

Now we can use the iteration patterns from lecture 6:

```{r}
# Define several CA coastal cities
locations <- data.frame(
  city = c("Santa Barbara", "Ventura", "San Luis Obispo"),
  lat  = c(34.42,           34.27,     35.28),
  lon  = c(-119.70,         -119.29,   -120.66)
)

# Use for loop to call our function for each row

for (i in 1:nrow(locations)) {
  city <- locations$city[i]
  lat  <- locations$lat[i]
  lon  <- locations$lon[i]
  
  aq_df <- get_pm25(lat, lon, past_days = 3)
  aq_df$city <- city
  
  if (i == 1) {
    aq_all <- aq_df
  } else {
    aq_all <- bind_rows(aq_all, aq_df)
  }
}


# Daily averages by city
aq_daily <- aq_all %>%
  mutate(date = as.Date(time)) %>%
  group_by(city, date) %>%
  summarize(mean_pm2_5 = mean(pm2_5, na.rm = TRUE), .groups = "drop")

ggplot(aq_daily, aes(x = date, y = mean_pm2_5, colour = city)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title  = "Daily Mean PM2.5 — Central Coast Cities",
    x      = "Date",
    y      = "PM2.5 (µg/m³)",
    colour = "City"
  ) +
  theme_minimal()
```

## EXTRA MATERIAL: MORE COMPLEX APIS, AND API WORKFLOW

Here’s another example from a different API and some additional on being
a good API citizen

## EXAMPLE 2: STREAMFLOW

Using the **USGS Water Services API** — the backbone of US hydrology
data

The structure of the data is more complex - you would probably need help
from the documentation to figure out how to navigate the nested list to
get to the actual values, and then parse those into a data frame

AI-is good support for this

But now you understand the basic work flow!

## USGS NATIONAL WATER INFORMATION SYSTEM

USGS maintains thousands of stream gauges across the US

API documentation: waterservices.usgs.gov

No API key needed — this is public data

You could again download data but API makes it easier

```{r}
# Get daily streamflow for the Santa Ynez River at Solvang (site 11128500)
req_usgs <- request("https://waterservices.usgs.gov/nwis/dv/") |>
  req_url_query(
    format        = "json",
    sites         = "11128500",    # USGS site number
    parameterCd   = "00060",       # parameter code for discharge (cfs)
    startDT       = "2024-01-01",
    endDT         = "2024-06-30"
  )

resp_usgs <- req_usgs |> req_perform()
flow_raw  <- resp_usgs |> resp_body_json()

# figure out what the data looks like
# USGS returns a very nested JSON structure — we need to navigate it to get to the actual values
# Website will tell you this
# you can also try names, str, and length to explore the structure of the list
names(flow_raw)  # top-level keys
```

```{r}
# Navigate the nested list to the actual values
flow_data <- flow_raw$value$timeSeries[[1]]$values[[1]]$value

# Parse into a data frame
# will use sapply to pull out date and value from each entry in the list
flow_df <- data.frame(
  date        = as.Date(substr(sapply(flow_data, `[[`, "dateTime"), 1, 10)),
  discharge   = as.numeric(sapply(flow_data, `[[`, "value"))
)

head(flow_df)
```

## VISUALISING STREAMFLOW

```{r}
ggplot(flow_df, aes(x = date, y = discharge)) +
  geom_area(fill = "steelblue", alpha = 0.4) +
  geom_line(colour = "steelblue") +
  scale_y_log10() +
  labs(
    title    = "Daily Streamflow — Santa Ynez River at Solvang",
    subtitle = "USGS Site 11128500 | Jan–Jun 2024",
    x        = "Date",
    y        = "Discharge (ft³/s, log scale)"
  ) +
  theme_minimal()
```

## ERROR HANDLING

APIs can fail — rate limits, bad parameters, network issues

Good practice: handle errors gracefully in your function

```{r}
get_pm25_safe <- function(lat, lon, past_days = 7) {
  # tryCatch lets your program keep running if one request fails
  tryCatch(
    {
      get_pm25(lat, lon, past_days)
    },
    error = function(e) {
      message("Request failed for (", lat, ", ", lon, "): ", e$message)
      return(NULL)   # return nothing for this location instead of crashing
    }
  )
}

# Now safe to loop over many locations
for (i in 1:nrow(locations)) {
  city <- locations$city[i]
  lat  <- locations$lat[i]
  lon  <- locations$lon[i]
  
  aq_df <- get_pm25_safe(lat, lon, past_days = 3)
  aq_df$city <- city
  
  if (i == 1) {
    aq_all <- aq_df
  } else {
    aq_all <- bind_rows(aq_all, aq_df)
  }
}
```

## BEING A GOOD API CITIZEN

APIs are a shared resource — follow these norms:

- **Read the docs** — understand rate limits before looping over
  hundreds of requests
- **Cache results** — save data locally so you don’t re-request what you
  already have
- **Throttle requests** — req_throttle() in httr2 request, or
  Sys.sleep() in loops
- **Use your key responsibly** — don’t share it or embed it in public
  code

```{r}
# Simple caching pattern
cache_file <- "data/aq_santa_barbara.csv"

if (file.exists(cache_file)) {
  aq_df <- read_csv(cache_file)
} else {
  aq_df <- get_pm25(34.42, -119.70, past_days = 30)
  write_csv(aq_df, cache_file)
}
```

## SUMMARY: THE API WORKFLOW

Every API interaction follows the same pattern:

1.  **Find the API** — check docs, understand endpoints and parameters
2.  **Build a request** — request() + req_url_query() + any auth headers
3.  **Send it** — req_perform()
4.  **Parse the response** — resp_body_json() → nested list → data frame
5.  **Clean and analyse** — tidyverse tools you already know
6.  **Wrap in a function** — for reuse and modularity

This connects directly to what we’ve been building all semester:

- Functions with clear contracts
- Iteration over inputs (map, for loops)
- Data wrangling with tidyverse

## USEFUL ENVIRONMENTAL APIS TO EXPLORE

| **API**                | **What it provides**        | **Key needed?** |
|:-----------------------|:----------------------------|:----------------|
| Open-Meteo             | Weather & air quality       | No              |
| USGS Water Services    | Streamflow, groundwater     | No              |
| iNaturalist            | Species observations        | No (read)       |
| NASA Earthdata / POWER | Climate, solar, precip      | Yes (free)      |
| NOAA CDO               | Historical climate records  | Yes (free)      |
| GBIF                   | Global biodiversity records | No              |
| EPA AQS                | US air quality monitoring   | Yes (free)      |
| OpenWeatherMap         | Current + forecast weather  | Yes (free tier) |

All offer free tiers suitable for research and coursework
