---
title: "datapractice"
format: html
---

```{r}
library(tidyverse)
# 
firestarts = runif(20, min=0, max=100)
ggplot(data.frame(firestarts=firestarts),
       aes(firestarts)) + geom_histogram()
```

```{r}
#
otterweight = rnorm(20, mean=10, sd=2)
ggplot(data.frame(otterweight),
       aes(otterweight)) + geom_histogram()
```
```{r}
#
pick = sample(seq(2000,2025), size=2, replace=T)
pick
#
pick = sample(seq(2000,2025), size=20, replace=F)
pick
```

```{r}
# practice
# write the R code to create test data where we are 
# Poisson distribution (rpois(lambda=1))
# 1000 samples, how many encounters?

set.seed(42)
daily_encounters <- rpois(n=30, lambda = 1)
daily_encounters
```
```{r}
# 1. Simulate 30 days of bear encounters
set.seed(42)
daily_encounters <- rpois(n = 30, lambda = 1)
total_encounters <- sum(daily_encounters)

# 2. Randomly assign those encounters to 1,000 visitors
visitor_ids <- sample(1:1000, size = total_encounters, replace = TRUE)

# 3. Find the maximum encounters for any one visitor
max(table(visitor_ids))

```

```{r}
data()
#
#
data(airquality)
#
help(airquality)
#
str(airquality)

```
```{r}
attributes(airquality)
```
```{r}
ggplot(airquality, aes(x=Temp, y=Ozone)) + geom_point()+ggtitle("Ozone vs. Temp")
```

```{r}
ozone_sample = sample(airquality$Ozone, size=5, replace=F)
ozone_sample
```
```{r}
airquality_clean <- na.omit(airquality)

ozone_sample = sample(1:nrow(airquality_clean), size=5, replace=5)
ozone_sample
```
```{r}
airquality_clean[ozone_sample,]
```
```{r}
ggplot(airquality_clean, aes(x=Temp, y=Ozone)) + geom_point() +
  geom_point(data=data.frame(Temp=airquality_clean$Temp[ozone_sample],
                             Ozone=airquality_clean$Ozone[ozone_sample]), color ="red")
```
```{r}
# find median wind speed
# sample 10 wind speed measurements
# compute their median

wind_sample = sample(airquality_clean$Wind, size=10, replace=F)
wind_sample

wind = median(wind_sample)
summary(wind) # 7.15


wind2 = median(airquality_clean$Wind)
```

```{r}
# 1. Clean the data (remove rows with NA)
airquality_clean <- na.omit(airquality)

# 2. Calculate the TRUE median (Full Dataset)
wind_full_median <- median(airquality_clean$Wind)

# 3. Calculate the SAMPLE median (10 random measurements)
set.seed(123) # Optional: keeps the sample the same every time you run it
wind_sample <- sample(airquality_clean$Wind, size = 10, replace = FALSE)
wind_sample_median <- median(wind_sample)

# 4. View Results
cat("Full Dataset Median:", wind_full_median, "\n")
cat("Sample Median (n=10):", wind_sample_median, "\n")

```


```{r}
#Factors
a = c(1.3, 1, 4, 1.3, 22)
mean(a)
```
```{r}
possible.fish = c("salmon","steelhead", "shark", "tuna", "cod")

catch1 = base::sample(possible.fish, size=20, replace=T)
catch1

summary(catch1)

catch1 = as.factor(catch1)

summary(catch1)

plot(catch1, col="blue")

mean(summary(catch1))

max(summary(catch1))

which.max(summary(catch1))

names(which.max(catch1))
```

```{r}
---
title: "datapractice"
format: html
---

```{r}
library(tidyverse)
# 
firestarts = runif(20, min=0, max=100)
ggplot(data.frame(firestarts=firestarts),
       aes(firestarts)) + geom_histogram()
```

```{r}
#
otterweight = rnorm(20, mean=10, sd=2)
ggplot(data.frame(otterweight),
       aes(otterweight)) + geom_histogram()
```
```{r}
#
pick = sample(seq(2000,2025), size=2, replace=T)
pick
#
pick = sample(seq(2000,2025), size=20, replace=F)
pick
```

```{r}
# practice
# write the R code to create test data where we are 
# Poisson distribution (rpois(lambda=1))
# 1000 samples, how many encounters?

set.seed(42)
daily_encounters <- rpois(n=30, lambda = 1)
daily_encounters
```
```{r}
# 1. Simulate 30 days of bear encounters
set.seed(42)
daily_encounters <- rpois(n = 30, lambda = 1)
total_encounters <- sum(daily_encounters)

# 2. Randomly assign those encounters to 1,000 visitors
visitor_ids <- sample(1:1000, size = total_encounters, replace = TRUE)

# 3. Find the maximum encounters for any one visitor
max(table(visitor_ids))

```

```{r}
data()
#
#
data(airquality)
#
help(airquality)
#
str(airquality)

```
```{r}
attributes(airquality)
```
```{r}
ggplot(airquality, aes(x=Temp, y=Ozone)) + geom_point()+ggtitle("Ozone vs. Temp")
```

```{r}
ozone_sample = sample(airquality$Ozone, size=5, replace=F)
ozone_sample
```
```{r}
airquality_clean <- na.omit(airquality)

ozone_sample = sample(1:nrow(airquality_clean), size=5, replace=5)
ozone_sample
```
```{r}
airquality_clean[ozone_sample,]
```
```{r}
ggplot(airquality_clean, aes(x=Temp, y=Ozone)) + geom_point() +
  geom_point(data=data.frame(Temp=airquality_clean$Temp[ozone_sample],
                             Ozone=airquality_clean$Ozone[ozone_sample]), color ="red")
```
```{r}
# find median wind speed
# sample 10 wind speed measurements
# compute their median

#wind_sample = sample(airquality_clean$Wind, size=10, replace=F)
#wind_sample

#wind = median(wind_sample)
#summary(wind) # 7.15


#wind2 = median(airquality_clean$Wind)
```

```{r}
# 1. Clean the data (remove rows with NA)
airquality_clean <- na.omit(airquality)

# 2. Calculate the TRUE median (Full Dataset)
wind_full_median <- median(airquality_clean$Wind)

# 3. Calculate the SAMPLE median (10 random measurements)
set.seed(123) # Optional: keeps the sample the same every time you run it
wind_sample2 <- sample(airquality_clean$Wind, size = 10, replace = FALSE)
wind_sample_median <- median(wind_sample2)

summary(wind_full_median)
summary(wind_sample_median)

```


```{r}
#Factors
a = c(1.3, 1, 4, 1.3, 22)
mean(a)
```
```{r}
possible.letters = c("a","c", "e", "g", "i")

letter1 = base::sample(possible.letters, size=10, replace=T)
letter1

summary(letter1)

letter1 = as.factor(letter1)

summary(letter1)

plot(letter1, col="blue")

mean(summary(letter1))

max(summary(letter1))

which.max(summary(letter1))

names(which.max(letter1))
```

```{r}
a = as.factor(c(1.3, 1, 4, 1.3, 22))
summary(a)

b = as.numeric(a)
summary(b)

```
```{r}
examplestring = "River Otters"
nchar(examplestring)
```
```{r}
sprintf("We found %d %s %3.1f miles from shore", 4, "Otters", 18.95632342)
```

