---
title: "functions2"
format: html
---

# FUNCTIONS

## PROGRAMMING ALWAYS STARTS…

{STYLE=“COLOR:ORANGE”}

- What is the goal
- How can you break this goal into sub-task
- For the overall project and each sub-task
  - What are the inputs
  - What are the outputs

## MODULARITY

Programs are made up of building blocks

When you start a programming/data science project - you should begin by:

- Breaking your goal down into individual pieces
- Identifying instructions that can be reused
- Identify tasks that might be useful to others
- Identify tasks that you might want to be able to substitute different methods for

## QUESTIONS ABOUT FUNCTIONS SO FAR

Homework

Lets generate a function that makes use of categorical data species diversity is a good example

Diversity (D) measures the probability that two individuals randomly selected from a sample will belong to the same species

To get value is between 0 and 1, with lower values associated with lower diversity 1-D from Assignment (Simpson Index)

## MY VERSION

```{r}
library(here)
library(tidyverse)

source(here("Week5/compute_diversity.R"))
compute_diversity
```

```{r}
possible.fish = as.factor(c("salmon","steelhead","shark","tuna","cod"))
# simulate a random recording of catch by fisherman

# we can control sample probablities as long as it sums to 1 - here we have a mix of fish but cod is most common

# to do this we assign probabities to each fish 
# high prob more likely to be sampled
# probabilities must sum to 1 
catch1 = sample(possible.fish, size=10, prob = c(0.2, 0.2, 0.1, 0.1, 0.4), replace=T)

# lets create a test case that should have low diversity, making most fish salmon
catch2 = sample(possible.fish, size=10, prob = c(0.8, 0.05, 0.05, 0.05, 0.05), replace=T)

# note that the diversity index works 'backwards' - higher values are less diverse 
compute_diversity(catch1)

compute_diversity(catch2)
```

## RETURNING VALUES FROM A FUNCTION

return(value) - but what if you want to return multiple items

Diversity and most frequent species?

In R -We use Lists

## LISTS

- Lists are the most “informal” data structures in R
- List are really useful for keeping track of and organizing groups of things that are not all the same
- A list could be a table where number of rows is different for each column
- A list can have numeric, character, factors all mixed together
- List are often used for returning more complex information from function (e.g. lm)

## EXAMPLE OF LISTS

```{r}
# make a list

sale = list(id=2, quality="high", contents=c("apple", "cherry"), cost=c(4,5))

sale
```

```{r}
# ways to access elements

sale$id

sale$what
```

```{r}
# you can also access by list item number
# the [x] denotes the xth item in the list

sale[[3]]

sale[["contents"]]

sale$contents[2]

sale[[3]][2]
```

```{r}
# how do you get the second element in the vector that has the contents
# there are two ways

# add to a list
sale$location = "Farmers market"
sale
```

```{r}
# or remove
sale$location = NULL
sale
```

```{r}
# some tricky things
# correct accessing items in list
sale$cost

sale[[4]]
```

```{r}
# works but
# sale[4]

sum(sale$cost)

sum(sale[[4]])
```

## FOR NOW - KEY PARTS OF LIST

- Lists can contain different types of data (numeric, character, factors, vectors, data frames, other lists)
- In R to create a list mylist = list(name=“Bob”, age=30, hobbies=c(“fishing”,“hiking”), scores=data.frame(test1=c(90,80), test2=c(85,95)))
- to access something in a list - two options mylist$name mylist[[“name”]]

There is more to say about lists but this is all we need for now

## USING LISTS TO RETURN MULTIPLE ITEMS FROM A FUNCTION

We can return a list from a function to give the user multiple pieces of information

```{r}
# a function that returns the max and min of a vector of numbers
minmax = function(x) {
  minx = min(x)
  maxx = max(x)
  return(list(min=minx, max=maxx))
}

minmax(c(3,5,2,8,1))
```

## MORE COMPLEX EXAMPLE

Lets say we want to compute a water quality index that is based on multiple measurements (e.g. temperature, dissolved oxygen, turbidity)

- check if each measurement is above, within, or below a threshold range
- return rating for each measurement (e.g. good, moderate, poor if above, within, or below threshold)
- return an overall numeric rating
- return an overall index of water quality (e.g. good, moderate, poor)

Look through function - and ask you neighbor any questions

```{r}
# Source the function
source(here("Week5/analyze_water_quality.R"))

# --- Use defaults (most common case) ---
results <- analyze_water_quality(ph = 7.2, dissolved_oxygen = 8.5, turbidity = 2.1)
results
```

```{r}
# --- Override one limit, e.g. stricter turbidity standard for drinking water ---
results_strict <- analyze_water_quality(
  ph = 7.2, dissolved_oxygen = 8.5, turbidity = 2.1,
  turbidity_max = 1.0
)
results_strict
```

## MY ANSWER

```{r}
library(tidyverse)
# repeat with a list
source(here("Week5/summarize_diversity.R"))

# lets compare two catches - with different levels of diversity
possible.fish =c("salmon","steelhead","shark","tuna","cod")

# we can control sample probablities as long as it sums to 1 - here we have a mix of fish but cod is most common

# to do this we assign probabities to each fish 
# high prob more likely to be sampled
# probabilities must sum to 1 
catch1 = sample(possible.fish, size=10, prob = c(0.2, 0.2, 0.1, 0.1, 0.4), replace=T)

# lets create a test case that should have low diversity, making most fish steelhead
catch2 = sample(possible.fish, size=10, prob = c(0.1, 0.7, 0.1, 0.1, 0.1), replace=T)

summarize_diversity
```

```{r}
# run function and save results to a variable
d1 = summarize_diversity(catch1)
d2 = summarize_diversity(catch2)
d1

d2

# just index
d1$simpson

d2$simpson
```

```{r}
source(here("Week5/practice.R"))

possible.tree = c("oak", "maple", "pine", "cherry")

forest1 = sample(possible.tree, size = 15, prob = c(0.2, 0.2, 0.2, 0.4), replace=T)

summarize.diversity

d1 = summarize.diversity(forest1)

d1

d1$simpson

```

```{r}
timber = list(id = 4, quality = "high", contents = possible.tree)

timber

forest2 = sample(possible.tree, size = 15, prob = c(0.1, 0.1, 0.6, 0.2), replace=T)

summarize.diversity

d2 = summarize.diversity(forest2)

d2

d2$simpson
```

## OTHER WAYS TO USE FUNCTIONS YOU MIGHT NOT THINK OF

- Use parameters to allow the user to make choices about what to return
- Use parameters to allow the user to make choices about how to compute something
- Use parameters to allow the user to make choices about what data to use
- Use parameters to apply different weighting to an index calculation

## EXAMPLE

Lets let user choose between two different ways to compute the index

and also whether to show a plot of the data or not

```{r}
library(here)
library(tidyverse)
source(here("Week5/summarize_diversity2.R"))

testfish = c("cod","trout","salmon")
# create a test catch that is dominated by cod, but has some trout and salmon
testcatch = sample(testfish, size=20, replace=T, prob=c(0.8,0.2,0.2))

# lets compare different ways to compute the index  
d1 = summarize_diversity(testcatch, index="A")
d2 = summarize_diversity(testcatch, index="B")

# just look at index
d1$simpson

d2$simpson

# run with instruction to plot
d3 = summarize_diversity(testcatch, index="B", showplot=TRUE)
d3$plt
```

